Declarative Probabilistic Logic Programming\\in Discrete-Continuous Domains

### 2024-09-10 202409_aij_dc_problog_aij_second/rebuttal.tex

\item Definition 7.4 states that $e^\oplus = (0, 0)$ is the neutral element for $\oplus$, So I
expect that $(r, n) \oplus  e^\oplus  = e^\oplus  \oplus  (r, n) = (r, n)$, But according to Definition
7.2 when $n > 0$, $(r, n) \oplus  e^\oplus  = e^\oplus  \oplus  (r, n) = e^\oplus $ . From this definition it
looks like a neutral element for $\oplus $ is $(0, \infty)$ but you have to include $\infty$ to
the set $\mathbb{Z}$. However, this implies that the inverse w.r.t. the $\otimes$ does not
exist. I have probably missed something here, it looks like all the following
relies on the fact that this structure is a commutative semiring. \citet{jacobs2021paradoxes} introduces explicitly the operation of difference and inverse, without
considering a neutral elements. Perhaps, the fact that is a semiring is not
so important for the paper.

\fixedrebuttal{indeed this was a mistake and the neutral element should have been defined as $e^\oplus =(0, \infty)$. We have fixed this in the manuscript. Note, however, that this has no serious consequences on any other part of the paper, except minor for the inverse elements. Specifically, we strike (as the reviewer suggested the definition of the inverse elements and define immediately subtraction and division \cf Definition~\ref{def:subdiv}.)
}

\commentrebuttal{Having no explicit inverse for $\otimes$ and $\oplus$ is not a problem: semirings do not necessitate inverse elements (\cf Definition~\ref{def:comm_semiring}).}

\item The solution to the problem of conditioning with $0$-probability events are
solved only partially. I.e., only in the case in which the $0$ probable events
are explicitly stated in an atom by using the primitive delta interval.
However, there are many other situations of events that are still possible
but with $0$ probability. As a simple example, the conjunction of the two
constraints $v \leq 1$ and $v \geq 1$ is equivalent to the constraint $v = 1$, and
I think that in a pure declarative approach, like the one that is pursued
by ProbLog they should have the same semantics. While the latter is
interpreted in terms of infinitesimal number the latter is not. How would
the infinitesimal approach generalize to the other comparison atoms? A
possible solution to this problem can be obtained by rewriting $x \leq y$ into
$(x < y ) \overline{\vee} (x=y)$
˙ (where $ \overline{\vee}$ denotes the xor operator).

\commentrebuttal{Here we have to disagree with the following statement: "While the latter is
interpreted in terms of infinitesimal number the [former] is not". This is not correct. Neither of the two are interpreted as infinitesimal numbers!
We would like to stress here that $x=y$ is not interpreted as an infinitesimal interval but simply as an indicator function where the equality has to hold strictly. This means that the semantics align between $v \leq 1 \land v \geq 1$ and  $v = 1$ as they are equivalent. Note that both of these expressions are, however, not equivalent to $v \doteq 1$, which is the actual infinitesimal interval. We already discuss this issue in Example~\ref{ex:conditional_prob}.
}

\item I also think that the semantics based on infinitesimals introduces other
paradoxes, For instance, if $x \sim \mathcal{N} (0, 1)$ $P (x < 0 | x = 1)$ is equal to $(0, 0)$.
However, by infinitesimal semantics this is equal to $( \frac{2}{y}
, -1)$ where $y$ is
the value of $(0, 1)$ in 1 (which is different from $(0, 0)$).

\commentrebuttal{Indeed, following the definition of a conditional probability with a zero probability conditioning event (using a limit), gives us
$P (x < 0 | x \doteq 1)=0 = (0\epsilon^0) = (0,0)$. Note that here we use the "$\doteq$" notation to indicate that we condition on an infinitesimal interval.

However, using infinitesimal numbers does not yield $(0,-1)$. If we write the conditional probability in question using infinitesimal numbers we get the following result:
\begin{align}
    \frac{
        \sum_{i}^{|S|} \alpha_{SIALW}(\ive{x^{(i)}<0}) \otimes \alpha_{SIALW}(\ive{x^{(i)} \doteq 1})
    }
    {
        \sum_{i}^{|S|} \alpha_{SIALW}(\ive{x^{(i)} \doteq 1})
    }
\end{align}
Because $x^{(i)} \doteq 1$ denotes an observation, all the samples $x^{(i)}$ are forced to take the value $1$. Plugging this into above expression gives:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
    \label{eq:rev3idiot}
    &=
    \frac{
        \sum_{i}^{|S|} (0,1)    
    }{
        \sum_{i}^{|S|}  (p(1),1)
    }
    \\
    &=
    \frac{
        (0,1)
    }
    {
        (|S|  p(1),1)
    }
    \\
    &=
    (0,0)
    \label{eq:r3:misconception}
\end{align}
That means if we straightforwardly apply the correct expression (the one that we derived in the manuscript) to compute the conditional probability we do obtain the result we expect.
}

\item Theorem 7.11 and Proposition 7.14 are not well formulated because the
meaning of $(x, 0) \approx (y, n)$ when $n \neq 0$ is not defined. Consider the previous
example $P (x < 0 | x  1) = (0, 0)$, however it is possible that S contains
at least one sample $S^{(i)}$ such that $S^{(i)}_x = 1$. In this case, the summation
at the denominator will be some value $(y, 1)$ while at the numerator you
will have an empty summation, which I suppose it is equal to $(0, 0)$. The
$\frac{(0,0)}{(x,1)}$
is equal to $(0, -1)$.

\commentrebuttal{Let us consider the left-hand side of Equation~\ref{eq:rev3idiot} above:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
\end{align}
Instead of multiplying out the numerator we can also cancel out the $(p(1),1)$ expression to obtain:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
    &
    =
    \frac{
        (0,0) \otimes (p(1),1) \otimes \cancel{\sum_{i}^{|S|} (1,0)}  
    }
    {
        (p(1),1) \otimes  \cancel{\sum_{i}^{|S|} (1,0)}
    }
    \\
    &=
    \frac{
        (0,0) \otimes \cancel{(p(1),1) } 
    }
    {
        \cancel{(p(1),1) }
    }
    \\
    &=
    (0,0)
\end{align}

While here we show that the particular case (pointed out by the reviewer) does not cause any problems (if one correctly applies the defined algebra), we also prove this for the general case in Subsection~\ref{app:proof:alwapproximation}.
We conclude that the apparent misbehavior never arises when the definitions are applied correctly.

Also note that for the case that we do not condition on $x\doteq 1$ but on $x=1$, we almost surely compute the correct result. We state this explicitly in Proposition~\ref{prop:alw_consistency} already.
}

\end{enumerate}

\subsection{Minor Points}
\begin{enumerate}
    \setcounter{enumi}{18}
\item line 150 “on” → “one”
\\
\fixedrebuttal{Fixed.}
\item line 175 “mutually marginally independent” → “mutually independent”
why you use marginally.
\\
\fixedrebuttal{We removed "marginally."}

\item Definition 3.3: “regular ground term” → “regular ground atom”
\\
\commentrebuttal{No, these are indeed ground terms and not ground atoms.}

\item Paragraph 901 explains the order of sampling, and introduces the term
“ancestral sample”. This is confusing because every sample must be an
ancestral sample since you need the value of the parent variables in order
to sample a variable. I suggest not to use this term and to explain before
Property 7.1, that to obtain a sample for all the variables we have to
proceed according to the partial order on the variables.

\commentrebuttal{With using the term "ancestral sampling" we are stressing the fact we perform sampling using ancestral sampling. This is in order to distinguish the samples from other sampling strategies, such as Gibbs sampling or any other MCMC method. As this MCMC methods are ubiquitous in probabilistic programming we us "ancestral sample" to make the sample strategy explicit.}

\item line 510: $L > 5$ → $\Lambda \geq 5$
\\
\fixedrebuttal{Fixed.}
\item Definition 4.21. in the first bullet, you assume that only one random
variable occurs in a random term, however in general this is not the case.
\\
\commentrebuttal{We do not believe that we make this assumption. We actually explicitly state "a distribution term that involves exavtly $k$ different random terms". Note this definition has moved and is now called Definition~\ref{def:elim-dc}.}
\item Definition 4.28: Clarify what is $P^{DF,*}$.
\\
\fixedrebuttal{We now clearly state this at the end of Definitino~\ref{def:adfree-to-core}.}
\item Definition 7.9: revise English and “evaluate” → “denotes”. The defini-
tion is hard to read. The reference to literals and negated literals is a
bit confusing. For instance, what happens to the negation of the comparison atom $v =/= 3$? Using the notation $[[s_k = x]]$ could also improve
readability.
\\
\fixedrebuttal{We rewrote the Definiton~\ref{def:sample_labeling_function}, hopefully improving the readability. We also use the Iverson bracket $\ive{\cdot}$ as suggested.}
\item Lines 1029-1031 it seems that you compare DC problog with itself. Is the
system introduced in [Fierence et al. 2015] also called DC-ProbLog?
\\
\fixedrebuttal{The language introduced by \citep{fierens2015inference} is called \problogsty. This was a typo on our side, and we change "\dcproblogsty" to "\problogsty",}
\item Definition 7.24: “given is” → “given”
\\
\fixedrebuttal{Fixed.}
\item It’s not completely clear if $\neg(x =/= 2)$ should be considered a negative
literal, or $x =/= 2$ is a negative literal of the literal $x=2$.
\\
\commentrebuttal{We discuss the issue of how to interpret negation in detail in Appendix~\ref{sec:non-mixture-dc} and Appendix~\ref{sec:dcproblog-dc}. Especially as \dcproblogsty and \dcsty have different semantics for negation. In short: when distributional clauses are present $x =/= 2$ is not necessriöy equivalent to $\neg(x=:=2)$ otherwise they are equivalent.}
\item Paragraph 185: is very obscure at this point of the paper. I believe it
could be put in a proper related work section.
\\
\fixedrebuttal{We removed the discussion in this paragraph and give now forward pointers to the related work Section~\ref{sec:related} and to Appendix~\ref{sec:dcproblog-dc}.}
\item The definition of infinitesimal interval is not very intuitive. In order to
grasp it I had to read the paper \citep{jacobs2021paradoxes}. I suggest reporting his explanation provided below the definition 5.2 of that paper.
\\
\fixedrebuttal{We added a reference to \Citet[Section 5.2]{jacobs2021paradoxes} just before Definition~\ref{def:inf_number}.}
\item Bibliography format is not correct.

\fixedrebuttal{Fixed.}

\end{enumerate}

\subsection{Reccomendations}

\begin{enumerate}
    \setcounter{enumi}{32}

\item The paper is potentially interesting but I think it needs to be improved in
the readability, i.e., make it shorter by concentrating on the major points,
and optimizing the usage of examples. The technical part should also
be double-checked. I might be wrong with my comments, in such a case
the authors should better explain the technical part, or if I’m right the
authors have to fix and revise (if necessary) the proofs. Since the paper
does not contain major technical theoretical results, most of the results
are applications/simple generalization of existing results, the paper should
provide some experiments that show the potential of the approach. I think
that to address all the above observations the paper needs a major revision,
which is my suggestion.

\comment{We hope that the changes we made to the paper addresses the concerns raised by the reviewer. Specifically,
\begin{enumerate}
    \item we substantially reduced the section on syntactic sugar
    \item we corrected our mistake regarding the neutral elements for the addition in our semiring and adapted the manuscript accordingly,
    \item we added experimental evidence that the developed SIALW algorithm exhibits advantageous behavior when compares to similar algorithms in the presence of low probability events. 
\end{enumerate}

We would also argue that our technical contribution does not constitute a simple extension of existing ideas. This is exemplified by the fact that a series of papers \citep{kersting2000bayesian,gutmann2010extending,gutmann2011magic,azzolini2021semantics} have indeed tackled the problem of unifying discrete and continuous random variables for probabilistic logic programming but none have succeeded in linking these discrete-continuous languages to the inference method of knowledge compilation -- the de facto standard for exact inference in the discrete domain. We achieve this using our IALW semiring. While we do deploy proof techniques developed in other papers, considerable effort has gone into correctly formulating them for the discrete-continuous domain. We are of the strong opinion that our theoretical contributions are not applications but strong generalizations of existing work.
}
\end{enumerate}
